---
title: "Correlations"
author: "John Blischak"
date: 2017-08-14
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

Calculate pairwise correlations between single cells.

## Setup

```{r packages, message=FALSE}
library("cowplot")
library("dplyr")
library("edgeR")
library("ggplot2")
library("stringr")
library("tidyr")
theme_set(theme_cowplot())
library("Biobase") # has to be loaded last to use `combine`
```

Import data.

```{r import}
eset <- readRDS("../data/eset.rds")
eset
```

Remove samples with bad cell number or TRA-1-60.

```{r filter}
eset_quality <- eset[, eset$cell_number == 1 & eset$tra1.60]
dim(eset_quality)
```

Separate by source.

```{r separate-by-source}
eset_ce <- eset_quality[fData(eset_quality)$source == "C. elegans", ]
head(featureNames(eset_ce))
eset_dm <- eset_quality[fData(eset_quality)$source == "D. melanogaster", ]
head(featureNames(eset_dm))
eset_ercc <- eset_quality[fData(eset_quality)$source == "ERCC",
                          eset_quality$ERCC != "Not added"]
head(featureNames(eset_ercc))
eset_hs <- eset_quality[fData(eset_quality)$source == "H. sapiens", ]
head(featureNames(eset_hs))
```

Define a function for filtering by percentage of cells in which a gene is detected.

```{r function-present}
present <- function(x, percent = 0.50) mean(x > 0) >= percent
```

## ERCC

Remove zeros.

```{r remove-zeros-ercc}
eset_ercc_clean <- eset_ercc[rowSums(exprs(eset_ercc)) != 0, ]
dim(eset_ercc_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-ercc}
eset_ercc_clean <- eset_ercc_clean[apply(exprs(eset_ercc_clean), 1, present), ]
dim(eset_ercc_clean)
```

Convert to log2 counts per million.

```{r cpm-ercc}
mol_ercc_cpm <- cpm(exprs(eset_ercc_clean), log = TRUE)
```

Calculate pairwise correlations.

```{r cor-ercc}
cor_ercc <- cor(mol_ercc_cpm)
cor_ercc <- cor_ercc[upper.tri(cor_ercc)]
summary(cor_ercc)
```

## Drosophila

Remove zeros.

```{r remove-zeros-dm}
eset_dm_clean <- eset_dm[rowSums(exprs(eset_dm)) != 0, ]
dim(eset_dm_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm}
eset_dm_clean <- eset_dm_clean[apply(exprs(eset_dm_clean), 1, present), ]
dim(eset_dm_clean)
```

Convert to log2 counts per million.

```{r cpm-dm}
mol_dm_cpm <- cpm(exprs(eset_dm_clean), log = TRUE)
```

Calculate pairwise correlations.

```{r cor-dm}
cor_dm <- cor(mol_dm_cpm)
cor_dm <- cor_dm[upper.tri(cor_dm)]
summary(cor_dm)
```

## Drosophila - 5 pg

Select only samples that received 5 pg.

```{r dm-5000}
eset_dm_5pg <- eset_dm[, eset_dm$fly == 5000]
dim(eset_dm_5pg)
```

Remove zeros.

```{r remove-zeros-dm_5pg}
eset_dm_5pg_clean <- eset_dm_5pg[rowSums(exprs(eset_dm_5pg)) != 0, ]
dim(eset_dm_5pg_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm_5pg}
eset_dm_5pg_clean <- eset_dm_5pg_clean[apply(exprs(eset_dm_5pg_clean), 1, present), ]
dim(eset_dm_5pg_clean)
```

Convert to log2 counts per million.

```{r cpm-dm_5pg}
mol_dm_5pg_cpm <- cpm(exprs(eset_dm_5pg_clean), log = TRUE)
```

Calculate pairwise correlations.

```{r cor-dm_5pg}
cor_dm_5pg <- cor(mol_dm_5pg_cpm)
cor_dm_5pg <- cor_dm_5pg[upper.tri(cor_dm_5pg)]
summary(cor_dm_5pg)
```

## Drosophila - 50 pg

Select only samples that received 50 pg.

```{r dm-50000}
eset_dm_50pg <- eset_dm[, eset_dm$fly == 50000]
dim(eset_dm_50pg)
```

Remove zeros.

```{r remove-zeros-dm_50pg}
eset_dm_50pg_clean <- eset_dm_50pg[rowSums(exprs(eset_dm_50pg)) != 0, ]
dim(eset_dm_50pg_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm_50pg}
eset_dm_50pg_clean <- eset_dm_50pg_clean[apply(exprs(eset_dm_50pg_clean), 1, present), ]
dim(eset_dm_50pg_clean)
```

Convert to log2 counts per million.

```{r cpm-dm_50pg}
mol_dm_50pg_cpm <- cpm(exprs(eset_dm_50pg_clean), log = TRUE)
```

Calculate pairwise correlations.

```{r cor-dm_50pg}
cor_dm_50pg <- cor(mol_dm_50pg_cpm)
cor_dm_50pg <- cor_dm_50pg[upper.tri(cor_dm_50pg)]
summary(cor_dm_50pg)
```

## C. elegans

Remove zeros.

```{r remove-zeros-ce}
eset_ce_clean <- eset_ce[rowSums(exprs(eset_ce)) != 0, ]
dim(eset_ce_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-ce}
eset_ce_clean <- eset_ce_clean[apply(exprs(eset_ce_clean), 1, present), ]
dim(eset_ce_clean)
```

Convert to log2 counts per million.

```{r cpm-ce}
mol_ce_cpm <- cpm(exprs(eset_ce_clean), log = TRUE)
# Remove samples with no observations for this subset
zeros_ce <- colSums(exprs(eset_ce_clean)) > 0
eset_ce_clean <- eset_ce_clean[, zeros_ce]
mol_ce_cpm <- mol_ce_cpm[, zeros_ce]
```

Calculate pairwise correlations.

```{r cor-ce}
cor_ce <- cor(mol_ce_cpm)
cor_ce <- cor_ce[upper.tri(cor_ce)]
summary(cor_ce)
```

## Human

Remove zeros.

```{r remove-zeros-hs}
eset_hs_clean <- eset_hs[rowSums(exprs(eset_hs)) != 0, ]
dim(eset_hs_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-hs}
eset_hs_clean <- eset_hs_clean[apply(exprs(eset_hs_clean), 1, present), ]
dim(eset_hs_clean)
```

Convert to log2 counts per million.

```{r cpm-hs}
mol_hs_cpm <- cpm(exprs(eset_hs_clean), log = TRUE)
```

Calculate pairwise correlations.

```{r cor-hs}
cor_hs <- cor(mol_hs_cpm)
cor_hs <- cor_hs[upper.tri(cor_hs)]
summary(cor_hs)
```

## Summary plot

Below are the pairwise correlations for each source of RNA. The numbers along
the bottom are the number of pairwise correlations included in each boxplot (not
all samples received the same spike-ins).

```{r pairwise-correlations}
boxplot(cor_ercc, cor_hs, cor_dm_50pg, cor_dm, cor_dm_5pg, cor_ce,
        names = c("ERCC", "Human", "Fly (50 pg)", "Fly (all)",
                  "Fly (5 pg)", "worm"),
        xlab = "Source", ylab = "Pearson correlation",
        main = "Pairwise correlations")
text(x = 1:6, y = -1, labels = sapply(list(cor_ercc, cor_hs, cor_dm_50pg, cor_dm, cor_dm_5pg, cor_ce), length))
```

## Pairwise correlations per C1 chip

Do the pairwise correlations change when looking within a chip versus between chips?

```{r calc_cor_pairs}
# Calculate pairwise correlations
# 
# x - matrix produced by edger::cpm(). Column names should be of the form
# 03162017_A01
# 
# id - optional character vector column
#
# method - a valid method to pass to cor()
calc_cor_pairs <- function(x, id = NULL, method = "pearson") {
  stopifnot(is.matrix(x),
            is.null(id) || (is.character(id) && length(id) == 1),
            method %in% c("pearson", "kendall", "spearman"))
  result <- x %>%
    cor(method = method) %>%
    as.data.frame %>%
    mutate(sample1 = colnames(x)) %>%
    gather(key = "sample2", value = "r", -sample1) %>%
    mutate(sample2 = as.character(sample2)) %>%
    filter(sample1 < sample2) %>% # only keep 1 of 2 duplicate entries
    arrange(sample1, sample2) %>%
    extract(col = sample1, into = c("experiment1", "well1"),
            regex = "([[:digit:]]+)-([[:alnum:]]+)") %>%
    extract(col = sample2, into = c("experiment2", "well2"),
            regex = "([[:digit:]]+)-([[:alnum:]]+)")
  n <- ncol(x)
  stopifnot(result$r < 1,
            nrow(result) == (n * n - n) / 2)
  if (!is.null(id)) {
    result$id <- id
  }
  return(result)
}
```

Cacluate pairwise correlations for each feature type and then combine them.

```{r cor-long}
cor_long_dm <- calc_cor_pairs(mol_dm_cpm, id = "Fly")
cor_long_dm_5pg <- calc_cor_pairs(mol_dm_5pg_cpm, id = "Fly 5 pg")
cor_long_dm_50pg <- calc_cor_pairs(mol_dm_50pg_cpm, id = "Fly 50 pg")
cor_long_ercc <- calc_cor_pairs(mol_ercc_cpm, id = "ERCC")
cor_long_hs <- calc_cor_pairs(mol_hs_cpm, id = "Human")
cor_long <- rbind(cor_long_dm, cor_long_dm_5pg, cor_long_dm_50pg,
                  cor_long_ercc, cor_long_hs)
```

The pairwise correlations are only slightly higher within a chip versus between a chip.

```{r cor-within-v-between}
cor_within_v_between <- cor_long %>%
  mutate(within = ifelse(experiment1 == experiment2,
                         "Within", "Between"))
ggplot(cor_within_v_between, aes(x = within, y = r)) +
  geom_boxplot() +
  facet_wrap(~id) +
  labs(title = "Pairwise correlations within versus between chips",
       x = "", y = "Correlation")
```

The pairwise correlations for a given RNA input source is similar across chips.

```{r cor-per-chip}
cor_per_chip <- cor_long %>%
  filter(experiment1 == experiment2)
ggplot(cor_per_chip, aes(x = id, y = r)) +
  geom_boxplot() +
  facet_wrap(~ experiment1) +
  labs(title = "Pairwise correlations per C1 chip",
       x = "", y = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

Note that the boxplots for Fly and Fly 5 or 50 pg are not identical even though
each chip has only 1 because these were processed separately above when
filtering genes which are present in at least 50% of samples.

## Chip effect versus individual effect

```{r prepare-chip-v-individual}
anno <- pData(eset)
head(cor_long_hs)
cor_hs_long_cat <- cor_long_hs %>%
  mutate(chip_id1 = anno[paste(experiment1, well1, sep = "-"), "chip_id"],
         chip_id2 = anno[paste(experiment2, well2, sep = "-"), "chip_id"],
         same_chip = experiment1 == experiment2,
         same_ind = chip_id1 == chip_id2,
         same_chip_same_ind = same_chip & same_ind,
         same_chip_diff_ind = same_chip & !same_ind,
         diff_chip_same_ind = !same_chip & same_ind,
         diff_chip_diff_ind = !same_chip & !same_ind) %>%
  gather("category", "assigned",
         same_chip_same_ind:diff_chip_diff_ind) %>%
  filter(assigned)
```

The effect of individual is stronger than the effect of chip.

```{r chip-v-individual}
cor_hs_long_cat$category <- factor(cor_hs_long_cat$category,
                                   levels = c("same_chip_same_ind",
                                              "diff_chip_same_ind",
                                              "same_chip_diff_ind",
                                              "diff_chip_diff_ind"),
                                   labels = c("Same chip\nSame ind",
                                              "Diff chip\nSame ind",
                                              "Same chip\nDiff ind",
                                              "Diff chip\nDiff ind"))
ggplot(cor_hs_long_cat, aes(x = category, y = r)) +
  geom_boxplot() +
  labs(title = "Chip effect v. individual effect",
       x = "", y = "Pairwise Pearson correlations")
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
