---
title: "Data collection overview"
author: "John Blischak"
date: 2017-10-25
output:
  html_document:
    toc: false
---

<!-- The file analysis/chunks.R contains chunks that define defau
lt settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
knitr::opts_chunk$set(echo = FALSE)
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r analysis, message=FALSE}
library("dplyr")
library("DT")
library("Biobase")
source("../code/functions.R")
eset <- readRDS("../data/eset.rds")
# Filter cells
quality <- read.table("../data/quality-single-cells.txt",
                      stringsAsFactors = FALSE)
colnames(quality) <- c("sample", "quality")
eset <- eset[, quality$quality]
# Filter genes
eset <- eset[fData(eset)$source %in% c("H. sapiens", "ERCC") , ]
genes <- read.table("../data/genes-pass-filter.txt", stringsAsFactors = FALSE)
colnames(genes) <- c("gene", "passed")
eset <- eset[genes$passed, ]
# Total molecules of human genes and ERCC spike-ins
sums <- colSums(exprs(eset))
collection <- pData(eset) %>%
  select(chip_id, detect_hs) %>%
  cbind(sums) %>%
  group_by(chip_id) %>%
  summarize(n = n(),
            # Round the medians to reduce the noise from tie breaks which cause
            # a trivial 0.5 to be added
            detect_med = round(median(detect_hs)),
            sum_med = round(median(sums))) %>%
  filter(n >= 5) %>%
  arrange(desc(n))
```

We have collected a total of `r sum(collection$n)` cells from a total of
`r nrow(collection)` individuals.

Column descriptions:

* **Individual** - This includes any individual that had at least 5 cells that
passed the [quality control cutoffs](sampleqc.html)

* **Number of cells** - Total number of cells for that individual that
passed the [quality control cutoffs](sampleqc.html)

* **Median number of genes detected** - This only includes human genes and ERCC
spike-ins which passed the [gene filtering cutoffs](filter-gene.html)

* **Median number of molecules** - This only includes the molecules that map to
human genes and ERCC spike-ins which passed the [gene filtering
cutoffs](filter-gene.html)

```{r datatable}
datatable(collection, options = list(pageLength = nrow(collection)),
          colnames = c("Individual", "Number of cells",
                       "Median number of genes detected",
                       "Median number of molecules"))
```
