---
title: "Filter genes"
author: "Po-Yuan Tung"
date: 2017-01-26
output: workflowr::wflow_html
---









## Setup

```{r packages, message=FALSE}
library("cowplot")
library("dplyr")
library("edgeR")
library("ggplot2")
library("knitr")
theme_set(theme_cowplot())
source("../code/functions.R")
library("Biobase") # has to be loaded last to use `combine`
```

Import data.

```{r import}
eset <- readRDS("../data/eset.rds")
dim(eset)
```

## Filter

### Low QC sample

We performed [quality control](sampleqc.html) to identify low quality single cells. Only keep high-quality single cells.

```{r quality}
quality <- read.table("../data/quality-single-cells.txt", stringsAsFactors = FALSE)
colnames(quality) <- c("sample", "quality")
eset <- eset[, quality$quality]
dim(eset)
```

### Over-expressed gene

We would like to remove genes which have greater than 4096 (4*4*4*4*4*4) molecules in any given single cell. These are above our theoretical maximum number of 6-bp UMIs (it can happen when a highly expressed gene as multiple start sites). We did not find any genes with molecules greater than 4096. We found r length(overexpressed_genes_1024) mitochondrial genes which have molecules larger than 1024

```{r human-genes}
## look at human genes
eset_hs <- eset[fData(eset)$source == "H. sapiens", ]
dim(eset_hs)
head(featureNames(eset_hs))

## identify gene with molecule > 4096
overexpressed_rows <- apply(exprs(eset_hs), 1, function(x) any(x >= 4096))
overexpressed_genes <- rownames(exprs(eset_hs))[overexpressed_rows]
overexpressed_genes

## identify gene with molecule > 1024
overexpressed_rows_1024 <- apply(exprs(eset_hs), 1, function(x) any(x >= 1024))
overexpressed_genes_1024 <- rownames(exprs(eset_hs))[overexpressed_rows]
overexpressed_genes_1024
```

### Lowly expressed 

We identify the lower cutoff using the mean log2 molecule counts per million (cpm) in the high quality single cells.

```{r cpm}
## convert to log2 cpm
mol_hs_cpm <- cpm(exprs(eset_hs), log = TRUE)
mol_hs_cpm_means <- rowMeans(mol_hs_cpm)
summary(mol_hs_cpm_means)
hist(mol_hs_cpm_means, xlab = "Mean log2 molecule cpm in single cell samples",
     ylab = "Number of genes", main = "Identifying expression cutoff")

## set a cutoff
lower_exp_cutoff <- 3
abline(v = lower_exp_cutoff, col = "red")
genes_pass_filter <- rownames(mol_hs_cpm)[mol_hs_cpm_means > lower_exp_cutoff]
```

`r length(genes_pass_filter)` genes have a mean log2 molecule cpm greater than `r lower_exp_cutoff`

```{r output}
mol_hs_cpm <- data.frame(mol_hs_cpm)
mol_hs_cpm$mean <- apply(mol_hs_cpm, 1, mean)
mol_hs_cpm$filter <- mol_hs_cpm$mean > lower_exp_cutoff
write.table(data.frame(row.names(mol_hs_cpm), mol_hs_cpm[,"filter"]),
            file = "../data/genes-pass-filter.txt", quote = FALSE,
            sep = "\t", row.names = FALSE, col.names = FALSE)
```



