---
title: "Mean and CV"
author: "Po-Yuan Tung"
date: 2018-03-17
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this analysis is to identify examples where mean across individuals are small but cv across individuals are large.

## Setup

```{r packages, message=FALSE}
library("cowplot")
library("dplyr")
library("edgeR")
library("ggplot2")
library("knitr")
theme_set(theme_cowplot())
source("../code/functions.R")
library("Biobase") # has to be loaded last to use `combine`
```

Import data.

```{r import}
eset <- readRDS("../data/eset.rds")
dim(eset)
```

Keep only human genes
```{r human-genes}
eset <- eset[fData(eset)$source == "H. sapiens" , ]
dim(eset) 
```

Only keep high-quality single cells.

```{r quality-cell}
quality <- read.table("../data/quality-single-cells.txt", stringsAsFactors = FALSE)
colnames(quality) <- c("sample", "quality")
eset <- eset[, quality$quality]
dim(eset)
```

Only keep genes that passed the filters

```{r gene}
genes <- read.table("../data/genes-pass-filter.txt", stringsAsFactors = FALSE)
colnames(genes) <- c("gene", "passed")
eset <- eset[genes$passed, ]

eset_data <- exprs(eset)
anno <- anno <- pData(eset)
```

## Mean and CV

First, we need to correct for collision probability and cpm standarization.
```{r collision}
eset_data_cr <- as.data.frame(-4^6*log(1-eset_data/4^6))
eset_data_cpm <- as.data.frame(cpm(eset_data_cr, log = TRUE))
dim(eset_data_cpm)
stopifnot(rownames(anno) == colnames(eset_data_cpm))
```

### Examples from SCDE
Look at the molecule counts of top overdispersion genes from [SCDE analysis](scde-all.html)

```{r scde}
## look at molecule count per individaul
anno$CDKN1A <- t(eset_data_cpm[row.names(eset_data_cpm) == "ENSG00000124762",])
plot_cv_CDKN1A <- ggplot(anno, 
       aes(x = factor(chip_id), y = CDKN1A)) +
       geom_boxplot() +
  labs(x = "Individual", y = "CDKN1A molecule count") +
  theme(axis.text.x = element_text(hjust=1, angle = 90))

plot_cv_CDKN1A

## look at mean and cv per individaul 

## create function CV 
cv <- function(x) {
  sd(x)/mean(x)
}

CDKN1A_in <- anno %>% 
  group_by(chip_id) %>% 
  summarise(individaul_cv = cv(CDKN1A),
            individaul_mean = mean.default(CDKN1A))

head(CDKN1A_in)
                            
ggplot(data.frame(CDKN1A_in), 
       aes(x = factor(chip_id), y = individaul_mean)) +
       geom_bar(stat = "identity") +
  labs(x = "Individual", y = "Mean") +
  theme(axis.text.x = element_text(hjust=1, angle = 90))

ggplot(data.frame(CDKN1A_in), 
       aes(x = factor(chip_id), y = individaul_cv)) +
       geom_bar(stat = "identity") +
  labs(x = "Individual", y = "CV") +
  theme(axis.text.x = element_text(hjust=1, angle = 90))
```

### Compute mean and cv per gene per individual
```{r mean-cv}
## create mean and cv matrix per individaul per gene
mean_matrix <- c()  
cv_matrix <- c()
  
for (ii in unique(anno$chip_id)) {
  ### extract index of current chip id
  index <- which(anno$chip_id == ii)
  ### restrict x1 to index column
  data_temp <- eset_data_cpm[, index]
  ### compute mean and var
  mean_matrix <- cbind(mean_matrix, apply(data_temp, 1, mean))
  cv_matrix <- cbind(cv_matrix, apply(data_temp, 1, cv))
}

## create funtion mean sruqe error
mse <- function(x) {
  sum((x - mean(x)) ^2)
} 

gene_mean <- apply(mean_matrix, 1, mse)
gene_cv <- apply(cv_matrix, 1, mse)

plot(log(gene_mean),
     log(gene_cv))
```

