---
title: "Partioning the variance across factors"
author: "John Blischak"
date: 2017-09-18
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

## Setup

```{r packages, message=FALSE}
library("cowplot")
library("edgeR")
library("ggplot2")
library("variancePartition")
theme_set(theme_cowplot())
library("Biobase") # has to be loaded last to use `combine`
```

Import data.

```{r import}
fname <- Sys.glob("../data/eset/*.rds")
eset <- Reduce(combine, Map(readRDS, fname))
eset
```

Remove samples with bad cell number, lack of TRA-1-60 signal, or an invalid
individual assignment.

```{r filter}
eset_quality <- eset[, eset$cell_number == 1 & eset$tra1.60 & eset$valid_id]
dim(eset_quality)
```

Separate by source.

```{r separate-by-source}
eset_ce <- eset_quality[fData(eset_quality)$source == "C. elegans", ]
head(featureNames(eset_ce))
eset_dm <- eset_quality[fData(eset_quality)$source == "D. melanogaster", ]
head(featureNames(eset_dm))
eset_ercc <- eset_quality[fData(eset_quality)$source == "ERCC",
                          eset_quality$ERCC != "Not added"]
head(featureNames(eset_ercc))
eset_hs <- eset_quality[fData(eset_quality)$source == "H. sapiens", ]
head(featureNames(eset_hs))
```

Define a function for filtering by percentage of cells in which a gene is detected.

```{r function-present}
present <- function(x, percent = 0.50) mean(x > 0) >= percent
```

Define a function for partitioning the variance.

```{r}
# https://bioconductor.org/packages/3.4/bioc/vignettes/variancePartition/inst/doc/variancePartition.R
calc_partition <- function(x, info, cores = 8) {
  stopifnot(c("experiment", "chip_id") %in% colnames(info))
  library("doParallel")
  cl <- makeCluster(cores)
  registerDoParallel(cl)
  form <- ~ (1|experiment) + (1|chip_id) 
  varPart <- fitExtractVarPartModel(x, form, info)
  vp <- sortCols(varPart)
  return(vp)
}
```

## ERCC

Remove zeros.

```{r remove-zeros-ercc}
eset_ercc_clean <- eset_ercc[rowSums(exprs(eset_ercc)) != 0, ]
dim(eset_ercc_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-ercc}
eset_ercc_clean <- eset_ercc_clean[apply(exprs(eset_ercc_clean), 1, present), ]
dim(eset_ercc_clean)
```

```{r cpm-ercc}
mol_ercc_cpm <- cpm(exprs(eset_ercc_clean), log = TRUE)
```

```{r partition-ercc}
part_ercc <- calc_partition(mol_ercc_cpm, pData(eset_ercc_clean))
plotPercentBars(part_ercc)
plotVarPart(part_ercc, main = sprintf("ERCC (n = %d)", nrow(part_ercc)))
```

## Drosophila

Remove zeros.

```{r remove-zeros-dm}
eset_dm_clean <- eset_dm[rowSums(exprs(eset_dm)) != 0, ]
dim(eset_dm_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm}
eset_dm_clean <- eset_dm_clean[apply(exprs(eset_dm_clean), 1, present), ]
dim(eset_dm_clean)
```

Convert to log2 counts per million.

```{r cpm-dm}
mol_dm_cpm <- cpm(exprs(eset_dm_clean), log = TRUE)
```

```{r partition-dm}
part_dm <- calc_partition(mol_dm_cpm, pData(eset_dm_clean))
plotPercentBars(part_dm)
plotVarPart(part_dm, main = sprintf("Fly (n = %d)", nrow(part_dm)))
```

## Drosophila - 5 pg

Select only samples that received 5 pg.

```{r dm-5000}
eset_dm_5pg <- eset_dm[, eset_dm$fly == 5000]
dim(eset_dm_5pg)
```

Remove zeros.

```{r remove-zeros-dm_5pg}
eset_dm_5pg_clean <- eset_dm_5pg[rowSums(exprs(eset_dm_5pg)) != 0, ]
dim(eset_dm_5pg_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm_5pg}
eset_dm_5pg_clean <- eset_dm_5pg_clean[apply(exprs(eset_dm_5pg_clean), 1, present), ]
dim(eset_dm_5pg_clean)
```

Convert to log2 counts per million.

```{r cpm-dm_5pg}
mol_dm_5pg_cpm <- cpm(exprs(eset_dm_5pg_clean), log = TRUE)
```

```{r partition-dm_5pg}
part_dm_5pg <- calc_partition(mol_dm_5pg_cpm, pData(eset_dm_5pg_clean))
plotPercentBars(part_dm_5pg)
plotVarPart(part_dm_5pg, main = sprintf("Fly (5 pg; n = %d)", nrow(part_dm_5pg)))
```

## Drosophila - 50 pg

Select only samples that received 50 pg.

```{r dm-50000}
eset_dm_50pg <- eset_dm[, eset_dm$fly == 50000]
dim(eset_dm_50pg)
```

Remove zeros.

```{r remove-zeros-dm_50pg}
eset_dm_50pg_clean <- eset_dm_50pg[rowSums(exprs(eset_dm_50pg)) != 0, ]
dim(eset_dm_50pg_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm_50pg}
eset_dm_50pg_clean <- eset_dm_50pg_clean[apply(exprs(eset_dm_50pg_clean), 1, present), ]
dim(eset_dm_50pg_clean)
```

Convert to log2 counts per million.

```{r cpm-dm_50pg}
mol_dm_50pg_cpm <- cpm(exprs(eset_dm_50pg_clean), log = TRUE)
```

```{r partition-dm_50pg}
part_dm_50pg <- calc_partition(mol_dm_50pg_cpm, pData(eset_dm_50pg_clean))
plotPercentBars(part_dm_50pg)
plotVarPart(part_dm_50pg, main = sprintf("Fly (50 pg; n = %d)", nrow(part_dm_50pg)))
```

## Human

Remove zeros.

```{r remove-zeros-hs}
eset_hs_clean <- eset_hs[rowSums(exprs(eset_hs)) != 0, ]
dim(eset_hs_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-hs}
eset_hs_clean <- eset_hs_clean[apply(exprs(eset_hs_clean), 1, present), ]
dim(eset_hs_clean)
```

Convert to log2 counts per million.

```{r cpm-hs}
mol_hs_cpm <- cpm(exprs(eset_hs_clean), log = TRUE)
```

```{r partition-hs}
part_hs <- calc_partition(mol_hs_cpm, pData(eset_hs_clean))
plotPercentBars(part_hs)
plotVarPart(part_hs, main = sprintf("Human (n = %d)", nrow(part_hs)))
```

## Final plot

```{r variance-partition}
x_labels <- scale_x_discrete(labels = c("C1 chip", "Individual", "Residuals"))
x_title <- labs(x = "Model coefficient")
plot_grid(plotVarPart(part_ercc, main = sprintf("ERCC (n = %d)", nrow(part_ercc))) +
            x_labels + x_title,
          plotVarPart(part_hs, main = sprintf("Human (n = %d)", nrow(part_hs))) +
            x_labels + x_title,
          labels = LETTERS[1:2])
```

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
