---
title: "PCA"
author: "John Blischak"
date: 2017-08-14
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

**Batch 1 only**

## Setup

```{r packages, message=FALSE}
library("cowplot")
library("dplyr")
library("edgeR")
library("ggplot2")
library("stringr")
library("tidyr")
theme_set(theme_cowplot())
source("../code/functions.R")
library("Biobase") # has to be loaded last to use `combine`
```

Import data.

```{r import}
eset <- readRDS("../data/eset.rds")
eset
```

Limit this analysis to batch 1 since it deals with the spike-ins.

```{r}
eset <- eset[, eset$batch == "b1"]
dim(eset)
```

Remove samples with bad cell number or TRA-1-60.

```{r filter}
eset_quality <- eset[, eset$cell_number == 1 & eset$tra1.60]
dim(eset_quality)
```

Separate by source.

```{r separate-by-source}
eset_ce <- eset_quality[fData(eset_quality)$source == "C. elegans", ]
head(featureNames(eset_ce))
eset_dm <- eset_quality[fData(eset_quality)$source == "D. melanogaster", ]
head(featureNames(eset_dm))
eset_ercc <- eset_quality[fData(eset_quality)$source == "ERCC",
                          eset_quality$ERCC != "Not added"]
head(featureNames(eset_ercc))
eset_hs <- eset_quality[fData(eset_quality)$source == "H. sapiens", ]
head(featureNames(eset_hs))
```

Define a function for filtering by percentage of cells in which a gene is detected.

```{r function-present}
present <- function(x, percent = 0.50) mean(x > 0) >= percent
```

## ERCC

Remove zeros.

```{r remove-zeros-ercc}
eset_ercc_clean <- eset_ercc[rowSums(exprs(eset_ercc)) != 0, ]
dim(eset_ercc_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-ercc}
eset_ercc_clean <- eset_ercc_clean[apply(exprs(eset_ercc_clean), 1, present), ]
dim(eset_ercc_clean)
```

```{r hist-ercc}
mol_ercc_cpm <- cpm(exprs(eset_ercc_clean), log = TRUE)
hist(rowMeans(mol_ercc_cpm))
```

```{r pca-ercc}
pca_ercc <- run_pca(mol_ercc_cpm)
plot_pca(pca_ercc$PCs, pcx = 1, pcy = 2, explained = pca_ercc$explained,
         metadata = pData(eset_ercc_clean),
         color = "experiment")
plot_pca(pca_ercc$PCs, pcx = 3, pcy = 4, explained = pca_ercc$explained,
         metadata = pData(eset_ercc_clean),
         color = "experiment")
plot_pca(pca_ercc$PCs, pcx = 5, pcy = 6, explained = pca_ercc$explained,
         metadata = pData(eset_ercc_clean),
         color = "experiment")
plot_pca(pca_ercc$PCs, pcx = 5, pcy = 6, explained = pca_ercc$explained,
         metadata = pData(eset_ercc_clean),
         color = "ERCC")
```

## Drosophila

Remove zeros.

```{r remove-zeros-dm}
eset_dm_clean <- eset_dm[rowSums(exprs(eset_dm)) != 0, ]
dim(eset_dm_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm}
eset_dm_clean <- eset_dm_clean[apply(exprs(eset_dm_clean), 1, present), ]
dim(eset_dm_clean)
```

Convert to log2 counts per million.

```{r hist-dm}
mol_dm_cpm <- cpm(exprs(eset_dm_clean), log = TRUE)
hist(rowMeans(mol_dm_cpm))
```

```{r pca-dm}
pca_dm <- run_pca(mol_dm_cpm)
plot_pca(pca_dm$PCs, pcx = 1, pcy = 2, explained = pca_dm$explained,
         metadata = pData(eset_dm_clean),
         color = "experiment")
plot_pca(pca_dm$PCs, pcx = 1, pcy = 2, explained = pca_dm$explained,
         metadata = pData(eset_dm_clean),
         color = "fly", factors = "fly")
```

## Drosophila - 5 pg

Select only samples that received 5 pg.

```{r dm-5000}
eset_dm_5pg <- eset_dm[, eset_dm$fly == 5000]
dim(eset_dm_5pg)
```

Remove zeros.

```{r remove-zeros-dm_5pg}
eset_dm_5pg_clean <- eset_dm_5pg[rowSums(exprs(eset_dm_5pg)) != 0, ]
dim(eset_dm_5pg_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm_5pg}
eset_dm_5pg_clean <- eset_dm_5pg_clean[apply(exprs(eset_dm_5pg_clean), 1, present), ]
dim(eset_dm_5pg_clean)
```

Convert to log2 counts per million.

```{r hist-dm-5pg}
mol_dm_cpm_5pg <- cpm(exprs(eset_dm_5pg_clean), log = TRUE)
hist(rowMeans(mol_dm_cpm_5pg))
```

```{r pca-dm-5pg}
pca_dm_5pg <- run_pca(mol_dm_cpm_5pg)
plot_pca(pca_dm_5pg$PCs, pcx = 1, pcy = 2, explained = pca_dm_5pg$explained,
         metadata = pData(eset_dm_5pg_clean),
         color = "experiment")
plot_pca(pca_dm_5pg$PCs, pcx = 3, pcy = 4, explained = pca_dm_5pg$explained,
         metadata = pData(eset_dm_5pg_clean),
         color = "experiment")
```

## Drosophila - 50 pg

Select only samples that received 50 pg.

```{r dm-50000}
eset_dm_50pg <- eset_dm[, eset_dm$fly == 50000]
dim(eset_dm_50pg)
```

Remove zeros.

```{r remove-zeros-dm_50pg}
eset_dm_50pg_clean <- eset_dm_50pg[rowSums(exprs(eset_dm_50pg)) != 0, ]
dim(eset_dm_50pg_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-dm_50pg}
eset_dm_50pg_clean <- eset_dm_50pg_clean[apply(exprs(eset_dm_50pg_clean), 1, present), ]
dim(eset_dm_50pg_clean)
```

Convert to log2 counts per million.

```{r hist-dm-50pg}
mol_dm_cpm_50pg <- cpm(exprs(eset_dm_50pg_clean), log = TRUE)
hist(rowMeans(mol_dm_cpm_50pg))
```

```{r pca-dm-50pg}
pca_dm_50pg <- run_pca(mol_dm_cpm_50pg)
plot_pca(pca_dm_50pg$PCs, pcx = 1, pcy = 2, explained = pca_dm_50pg$explained,
         metadata = pData(eset_dm_50pg_clean),
         color = "experiment")
plot_pca(pca_dm_50pg$PCs, pcx = 3, pcy = 4, explained = pca_dm_50pg$explained,
         metadata = pData(eset_dm_50pg_clean),
         color = "experiment")
plot_pca(pca_dm_50pg$PCs, pcx = 5, pcy = 6, explained = pca_dm_50pg$explained,
         metadata = pData(eset_dm_50pg_clean),
         color = "experiment")
```

## C. elegans

Remove zeros.

```{r remove-zeros-ce}
eset_ce_clean <- eset_ce[rowSums(exprs(eset_ce)) != 0, ]
dim(eset_ce_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-ce}
eset_ce_clean <- eset_ce_clean[apply(exprs(eset_ce_clean), 1, present), ]
dim(eset_ce_clean)
```

Convert to log2 counts per million.

```{r hist-ce}
mol_ce_cpm <- cpm(exprs(eset_ce_clean), log = TRUE)
# Remove samples with no observations for this subset
zeros_ce <- colSums(exprs(eset_ce_clean)) > 0
eset_ce_clean <- eset_ce_clean[, zeros_ce]
mol_ce_cpm <- mol_ce_cpm[, zeros_ce]
hist(rowMeans(mol_ce_cpm))
```

```{r pca-ce}
pca_ce <- run_pca(mol_ce_cpm)
plot_pca(pca_ce$PCs, pcx = 1, pcy = 2, explained = pca_ce$explained,
         metadata = pData(eset_ce_clean),
         color = "experiment")
plot_pca(pca_ce$PCs, pcx = 1, pcy = 2, explained = pca_ce$explained,
         metadata = pData(eset_ce_clean),
         color = "worm", factors = "worm")
```

## Human

Remove zeros.

```{r remove-zeros-hs}
eset_hs_clean <- eset_hs[rowSums(exprs(eset_hs)) != 0, ]
dim(eset_hs_clean)
```

Only keep genes which are observed in at least 50% of the samples.

```{r present-hs}
eset_hs_clean <- eset_hs_clean[apply(exprs(eset_hs_clean), 1, present), ]
dim(eset_hs_clean)
```

Convert to log2 counts per million.

```{r hist-hs}
mol_hs_cpm <- cpm(exprs(eset_hs_clean), log = TRUE)
hist(rowMeans(mol_hs_cpm))
```

```{r pca-hs}
pca_hs <- run_pca(mol_hs_cpm)
plot_pca(pca_hs$PCs, pcx = 1, pcy = 2, explained = pca_hs$explained,
         metadata = pData(eset_hs_clean), color = "experiment")
plot_pca(pca_hs$PCs, pcx = 3, pcy = 4, explained = pca_hs$explained,
         metadata = pData(eset_hs_clean), color = "experiment")
```

Visualizing how cells cluster by chip and individual for a subset of
individuals.

```{r pca-hs-individual}
pca_hs_data <- cbind(pca_hs$PCs[, 1:4], pData(eset_hs_clean))
hs_pc1v2 <- ggplot(pca_hs_data %>% filter(chip_id %in% c("NA18505",
                                                         "NA18507",
                                                         "NA18508")),
                   aes(x = PC1, y = PC2, color = experiment)) +
  geom_text(aes(label = str_sub(chip_id, 6, 7)))
hs_pc1v2
hs_pc3v4 <- hs_pc1v2 %+% aes(x = PC3, y = PC4)
hs_pc3v4
```

Visualizing C1 chips 04132017 and 04142017 because they share 3 individuals. PC1
and PC2 separate outlier cells. PC3 appears to separate by chip followed by
individual in PC4.

```{r pca-cross-chip}
pca_cross_chip_1v2 <- ggplot(pca_hs_data %>%
                               filter(experiment %in% c("04132017", "04142017"),
                                      chip_id %in% c("NA18498", "NA18510", "NA18520",
                                                     "NA18522", "NA19203")),
                             aes(x = PC1, y = PC2, shape = experiment,
                                 color = chip_id)) +
  geom_point() +
  labs(title = "C1 chips 04132017 & 04142017")
pca_cross_chip_3v4 <- pca_cross_chip_1v2 %+% aes(x = PC3, y = PC4)
plot_grid(pca_cross_chip_1v2, pca_cross_chip_3v4, labels = LETTERS[1:2])
```


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
