---
title: "PCA"
author: "John Blischak"
date: 2017-08-14
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

## Setup

```{r packages, message=FALSE}
library("Biobase")
library("cowplot")
library("dplyr")
library("edgeR")
library("ggplot2")
library("readr")
library("stringr")
library("tidyr")
theme_set(theme_cowplot())
source("../code/functions.R")
```

Import molecules.

```{r import-molecules, message=FALSE, cache=FALSE}
molecules <- read_tsv(gzfile("../data/molecules.txt.gz"))
dim(molecules)
molecules[1:5, 1:10]
colnames(molecules)[1:20]
```

Import annotation.

```{r read-anno, message=FALSE}
anno <- read_tsv("../data/batch1_qc.txt",
                 col_types = cols(ERCC = col_character()))
# Cleanup ERCC column
anno$ERCC <- ifelse(is.na(anno$ERCC), "Not added", anno$ERCC)
anno$ERCC <- factor(anno$ERCC, levels = c("Not added", "1:100000", "1:50000"),
                    labels = c("Not added", "100x dilution", "50x dilution"))
anno$ERCC <- as.character(anno$ERCC)
dim(anno)
colnames(anno)
```

Merge the data sets.

```{r merge}
mol_anno <- merge(molecules, anno, by = c("experiment", "well"))
stopifnot(nrow(mol_anno) == nrow(molecules))
```

Remove samples with bad cell number or TRA-1-60.

```{r filter}
mol_anno <- mol_anno %>%
  filter(cell_number == 1, tra1.60 == 1)
dim(mol_anno)
```

Separate by source.

```{r separate-by-source}
mol_ce <- mol_anno %>% select(experiment:lane, cell_number:index, starts_with("WBGene"))
mol_dm <- mol_anno %>% select(experiment:lane, cell_number:index, starts_with("FBGn"))
mol_ercc <- mol_anno %>% select(experiment:lane, cell_number:index, starts_with("ERCC"))
mol_hs <- mol_anno %>% select(experiment:lane, cell_number:index, starts_with("ENSG"))
```

## ERCC

Remove dashes in column names.

```{r dashes}
colnames(mol_ercc) <- str_replace(colnames(mol_ercc), "-", "")
```

Remove samples that did not receive ERCC spike-in and change the name.

```{r ercc-not-added}
mol_ercc <- mol_ercc %>% filter(ERCC != "Not added") %>% rename(annoERCC = ERCC)
```

Remove zeros.

```{r}
mol_ercc <- mol_ercc %>%
  select_if(function(col) is.character(col) || sum(col) > 0)
dim(mol_ercc)
```

Convert to matrix.

```{r}
mol_ercc_mat <- mol_ercc %>%
  select(starts_with("ERCC")) %>%
  t
```

Only keep ERCC which are observed in at least 50% of the samples.

```{r}
present <- function(x, percent = 0.50) mean(x > 0) >= percent
mol_ercc_mat <- mol_ercc_mat[apply(mol_ercc_mat, 1, present), ]
dim(mol_ercc_mat)
```

```{r hist-ercc}
mol_ercc_cpm <- cpm(mol_ercc_mat, log = TRUE)
hist(rowMeans(mol_ercc_cpm))
```

```{r pca-ercc}
pca_ercc <- run_pca(mol_ercc_cpm)
plot_pca(pca_ercc$PCs, pcx = 1, pcy = 2, explained = pca_ercc$explained,
         metadata = mol_ercc %>% select(experiment:index),
         color = "experiment")
plot_pca(pca_ercc$PCs, pcx = 3, pcy = 4, explained = pca_ercc$explained,
         metadata = mol_ercc %>% select(experiment:index),
         color = "experiment")
plot_pca(pca_ercc$PCs, pcx = 5, pcy = 6, explained = pca_ercc$explained,
         metadata = mol_ercc %>% select(experiment:index),
         color = "experiment")
plot_pca(pca_ercc$PCs, pcx = 5, pcy = 6, explained = pca_ercc$explained,
         metadata = mol_ercc %>% select(experiment:index),
         color = "annoERCC")
```

## Drosophila

Remove zeros.

```{r}
mol_dm <- mol_dm %>%
  select_if(function(col) is.character(col) || sum(col) > 0)
dim(mol_dm)
```

Convert to matrix.

```{r}
mol_dm_mat <- mol_dm %>%
  select(starts_with("FBgn")) %>%
  t
```

Only keep fly genes which are observed in at least 50% of the samples.

```{r}
mol_dm_mat <- mol_dm_mat[apply(mol_dm_mat, 1, present), ]
dim(mol_dm_mat)
```

```{r hist-dm}
mol_dm_cpm <- cpm(mol_dm_mat, log = TRUE)
hist(rowMeans(mol_dm_cpm))
```

```{r pca-dm}
pca_dm <- run_pca(mol_dm_cpm)
plot_pca(pca_dm$PCs, pcx = 1, pcy = 2, explained = pca_dm$explained,
         metadata = mol_dm %>% select(experiment:index),
         color = "experiment")
plot_pca(pca_dm$PCs, pcx = 1, pcy = 2, explained = pca_dm$explained,
         metadata = mol_dm %>% select(experiment:index),
         color = "fly", factors = "fly")
```

## C. elegans

Remove zeros.

```{r}
mol_ce <- mol_ce %>%
  select_if(function(col) is.character(col) || sum(col) > 0)
dim(mol_ce)
```

Convert to matrix.

```{r}
mol_ce_mat <- mol_ce %>%
  select(starts_with("WBGene")) %>%
  t
```

Only keep worm genes which are observed in at least 50% of the samples.

```{r}
mol_ce_mat <- mol_ce_mat[apply(mol_ce_mat, 1, present), ]
dim(mol_ce_mat)
```

```{r hist-ce}
mol_ce_cpm <- cpm(mol_ce_mat, log = TRUE)
# Remove samples with no observations for this subset
zeros_ce <- colSums(mol_ce_mat) > 0
mol_ce_cpm <- mol_ce_cpm[, zeros_ce]
hist(rowMeans(mol_ce_cpm))
```

```{r pca-ce}
pca_ce <- run_pca(mol_ce_cpm)
plot_pca(pca_ce$PCs, pcx = 1, pcy = 2, explained = pca_ce$explained,
         metadata = mol_ce %>% select(experiment:index) %>% filter(zeros_ce),
         color = "experiment")
plot_pca(pca_ce$PCs, pcx = 1, pcy = 2, explained = pca_ce$explained,
         metadata = mol_ce %>% select(experiment:index) %>% filter(zeros_ce),
         color = "worm", factors = "worm")
```

## Human

Remove zeros.

```{r}
mol_hs <- mol_hs %>%
  select_if(function(col) is.character(col) || sum(col) > 0)
dim(mol_hs)
```

Convert to matrix.

```{r}
mol_hs_mat <- mol_hs %>%
  select(starts_with("ENSG")) %>%
  t
```

Only keep human genes which are observed in at least 50% of the samples.

```{r}
mol_hs_mat <- mol_hs_mat[apply(mol_hs_mat, 1, present), ]
dim(mol_hs_mat)
```

```{r hist-hs}
mol_hs_cpm <- cpm(mol_hs_mat, log = TRUE)
hist(rowMeans(mol_hs_cpm))
```

```{r pca-hs}
pca_hs <- run_pca(mol_hs_cpm)
plot_pca(pca_hs$PCs, pcx = 1, pcy = 2, explained = pca_hs$explained,
         metadata = mol_hs %>% select(experiment:index), color = "experiment")
plot_pca(pca_hs$PCs, pcx = 3, pcy = 4, explained = pca_hs$explained,
         metadata = mol_hs %>% select(experiment:index), color = "experiment")
```

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
