---
title: "QC of single cell libraries"
author: "PoYuan Tung"
date: 2017-09-13
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

This is for qc of the samples. Based on obsevation under the scope and the sequencing results, samples with bad quality will be removed.

## Setup

```{r packages, message=FALSE}
library("cowplot")
library("dplyr")
library("ggplot2")
library("tidyr")
library("edgeR")
theme_set(theme_cowplot())
source("../code/functions.R")
library("Biobase")
```

```{r color}
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

Import data.

```{r import}
fname <- Sys.glob("../data/eset/*.rds")
eset <- Reduce(combine, Map(readRDS, fname))
anno <- pData(eset)
```

## Total mapped reads reads

```{r total-reads}
## calculate the cut-off  
cut_off_reads <- quantile(anno[anno$cell_number == 0,"mapped"], 0.95)

cut_off_reads

anno$cut_off_reads <- anno$mapped > cut_off_reads

## numbers of cells 
sum(anno[anno$cell_number == 1, "mapped"] > cut_off_reads)

sum(anno[anno$cell_number == 1, "mapped"] <= cut_off_reads)

## density plots
plot_reads <- ggplot(anno[anno$cell_number == 0 |
                          anno$cell_number == 1 , ],
       aes(x = mapped, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_reads, colour="grey", linetype = "longdash") +
       labs(x = "Total mapped reads", title = "Number of total mapped reads", fill = "Cell number")

plot_reads
```

## Unmapped ratios
Note: Using the 5% cutoff of samples with no cells excludes all the samples

```{r unmapped-ratios}
## calculate unmapped ratios
anno$unmapped_ratios <- anno$unmapped/anno$umi

## cut off 
cut_off_unmapped <- quantile(anno[anno$cell_number == 0,"unmapped_ratios"], 0.05)

cut_off_unmapped

anno$cut_off_unmapped <- anno$unmapped_ratios < cut_off_unmapped

## numbers of cells 
sum(anno[anno$cell_number == 1, "unmapped_ratios"] >= cut_off_unmapped)

sum(anno[anno$cell_number == 1, "unmapped_ratios"] < cut_off_unmapped)

## density plots
plot_unmapped <- ggplot(anno[anno$cell_number == 0 |
                             anno$cell_number == 1 , ],
       aes(x = unmapped_ratios *100, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_unmapped *100, colour="grey", linetype = "longdash") +
       labs(x = "Unmapped reads/ total reads", title = "Unmapped reads percentage")

plot_unmapped
```

## ERCC percentage 
Note: Beacuse not all samples include ERCC, this is not a good cutoff.

```{r ercc-percentage}
## calculate ercc reads percentage
anno$ercc_percentage <- anno$reads_ercc / anno$mapped

## cut off 
cut_off_ercc <- quantile(anno[anno$cell_number == 0,"ercc_percentage"], 0.05)

cut_off_ercc

anno$cut_off_ercc <- anno$ercc_percentage < cut_off_ercc

## numbers of cells 
sum(anno[anno$cell_number == 1, "ercc_percentage"] >= cut_off_ercc)

sum(anno[anno$cell_number == 1, "ercc_percentage"] < cut_off_ercc)

## density plots
plot_ercc <- ggplot(anno[anno$cell_number == 0 |
                                anno$cell_number == 1 , ],
       aes(x = ercc_percentage *100, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_ercc *100, colour="grey", linetype = "longdash") +
       labs(x = "ERCC reads / total mapped reads", title = "ERCC reads percentage")

plot_ercc
```

## Spike-in percentage 
Note: Using the percentage of all the kinds of spike-in as the cutoff. Instead of 5%, 10% seem to be more reasonable due to different amounts of total spike-in. 

```{r spike-percentage}
## calculate worm and fly reads percentage
anno$spike_percentage <- apply(anno[,18:20],1,sum) / anno$mapped

## cut off 
cut_off_spike <- quantile(anno[anno$cell_number == 0,"spike_percentage"], 0.10)

cut_off_spike

anno$cut_off_spike <- anno$spike_percentage < cut_off_spike

## numbers of cells 
sum(anno[anno$cell_number == 1, "spike_percentage"] >= cut_off_spike)

sum(anno[anno$cell_number == 1, "spike_percentage"] < cut_off_spike)

## density plots
plot_spike <- ggplot(anno[anno$cell_number == 0 |
                          anno$cell_number == 1 , ],
       aes(x = spike_percentage *100, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_spike *100, colour="grey", linetype = "longdash") +
       labs(x = "spike-in reads / total mapped reads", title = "Spike-in reads percentage")

plot_spike
```

## Number of genes detected 
```{r gene-number}
## cut off 
cut_off_genes <- quantile(anno[anno$cell_number == 0,"detect_hs"], 0.95)

cut_off_genes

anno$cut_off_genes <- anno$detect_hs > cut_off_genes

## numbers of cells 
sum(anno[anno$cell_number == 1, "detect_hs"] > cut_off_genes)

sum(anno[anno$cell_number == 1, "detect_hs"] <= cut_off_genes)

## density plots
plot_gene <- ggplot(anno[anno$cell_number == 0 |
                         anno$cell_number == 1 , ],
       aes(x = detect_hs, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_genes, colour="grey", linetype = "longdash") +
       labs(x = "Gene numbers", title = "Numbers of detected genes")

plot_gene
```

```{r density plots}
plot_grid(plot_reads + theme(legend.position=c(.7,.7)),
          plot_unmapped + theme(legend.position = "none"),
          plot_spike + theme(legend.position = "none"), 
          plot_gene + theme(legend.position = "none"),
          labels = LETTERS[1:4])
```

## Linear Discriminat Analysis
### Total molecule vs concentration 
```{r lda, fig.width=12, fig.height=6}
library(MASS)
## create 3 groups according to cell number
group_3 <- rep("two",dim(anno)[1])
         group_3[grep("0", anno$cell_number)] <- "no"
         group_3[grep("1", anno$cell_number)] <- "one"

## create data frame
data <- anno %>% dplyr::select(experiment:concentration, mapped, molecules)
data <- data.frame(data, group = group_3)

## perform lda
data_lda <- lda(group ~ concentration + molecules, data = data)
data_lda_p <- predict(data_lda, newdata = data[,c("concentration", "molecules")])$class

## determine how well the model fix
table(data_lda_p, data[, "group"])
data$data_lda_p <- data_lda_p


## plot before and after
plot_before <- ggplot(data, aes(x = concentration, y = molecules / 10^3,
               color = as.factor(group))) +
               geom_text(aes(label = cell_number, alpha = 0.5)) +
               labs(x = "Concentration", y = "Gene molecules (thousands)", title = "Before") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")


plot_after <- ggplot(data, aes(x = concentration, y = molecules / 10^3,
               color = as.factor(data_lda_p))) +
               geom_text(aes(label = cell_number, alpha = 0.5)) +
               labs(x = "Concentration", y = "Gene molecules (thousands)", title = "After") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")

plot_grid(plot_before + theme(legend.position=c(.8,.85)), 
          plot_after + theme(legend.position = "none"),
          labels = LETTERS[1:2])
```

### Reads to molecule conversion
```{r convertion, fig.width=12, fig.height=6}
## calculate convertion
anno$dm_conversion <- anno$mol_dm / anno$reads_dm

anno$conversion <- anno$mol_hs / anno$reads_hs

ggplot(anno, aes(x = dm_conversion, y = conversion,
  color = as.factor(cell_number))) +
  geom_text(aes(label = cell_number)) +
  labs(x = "Convertion of Fly spike-ins", y = "Conversion of genes") +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")

## try lda
data$conversion <- anno$conversion
data$dm_conversion <- anno$dm_conversion

data_ercc_lda <- lda(group ~ dm_conversion + conversion, data = data)

data_ercc_lda_p <- predict(data_ercc_lda,  newdata = data[,c("dm_conversion", "conversion")])$class

## determine how well the model fix
table(data_ercc_lda_p, data[, "group"])

data$data_ercc_lda_p <- data_ercc_lda_p


## cutoff
out_ercc_con <- anno %>% filter(cell_number == "1", dm_conversion > .094)

anno$conversion_outlier <- anno$cell_number == 1 & anno$dm_conversion > .094

## plot before and after
plot_ercc_before <- ggplot(data, aes(x = dm_conversion, y = conversion,
               color = as.factor(group))) +
               geom_text(aes(label = cell_number, alpha = 0.5)) +
               labs(x = "Convertion of Fly spike-ins", y = "Conversion of genes", title = "Before") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")

plot_ercc_after <- ggplot(data, aes(x = dm_conversion, y = conversion,
               color = as.factor(data_ercc_lda_p))) +
               geom_text(aes(label = cell_number, alpha = 0.5)) +
               labs(x = "Convertion of Fly spike-ins", y = "Conversion of genes", title = "After") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")

plot_grid(plot_ercc_before, 
          plot_ercc_after,
          labels = LETTERS[3:4])
```

## Filter

### Final list
```{r filter}
## all filter
anno$filter_all <- anno$cell_number == 1 &
                   anno$valid_id &
                   anno$cut_off_reads &
                ##   anno$cut_off_unmapped &
                ##   anno$cut_off_ercc &
                   anno$cut_off_spike &
                   anno$cut_off_genes 
sort(table(anno[anno$filter_all, "chip_id"]))
```

### Plots
```{r plots, fig.width=12, fig.heigh=12, dpi=300}

genes_unmapped <-  ggplot(anno,
                   aes(x = detect_hs, y = unmapped_ratios * 100,
                       col = as.factor(valid_id), 
                       label = as.character(cell_number),
                       height = 600, width = 2000)) +
                   scale_colour_manual(values=cbPalette) +
                   geom_text(fontface = 3, alpha = 0.5) + 
                   geom_vline(xintercept = cut_off_genes, 
                              colour="grey", linetype = "longdash") +
                   geom_hline(yintercept = cut_off_unmapped * 100, 
                              colour="grey", linetype = "longdash") +
                   labs(x = "Number of detected genes / sample", 
                        y = "Percentage of unmapped reads (%)") 

genes_spike <- ggplot(anno,
               aes(x = detect_hs, y = spike_percentage * 100,
                   col = as.factor(valid_id), 
                   label = as.character(cell_number), 
                   height = 600, width = 2000)) +
               scale_colour_manual(values=cbPalette) +
               scale_shape_manual(values=c(1:10)) +
               geom_text(fontface = 3, alpha = 0.5) + 
               geom_vline(xintercept = cut_off_genes, 
                          colour="grey", linetype = "longdash") +
               geom_hline(yintercept = cut_off_spike * 100, 
                          colour="grey", linetype = "longdash") +
               labs(x = "Number of detected genes / samlpe", 
                    y = "Percentage of spike-in reads (%)") 

reads_unmapped_num <-  ggplot(anno,
                       aes(x = mapped, y = unmapped_ratios * 100,
                           col = as.factor(chip_id), 
                           label = as.character(cell_number), 
                           height = 600, width = 2000)) +
                       geom_text(fontface = 3, alpha = 0.5) + 
                       geom_vline(xintercept = cut_off_reads, 
                                  colour="grey", linetype = "longdash") +
                       geom_hline(yintercept = cut_off_unmapped * 100,
                                  colour="grey", linetype = "longdash") +
                       labs(x = "Total mapped reads / sample", 
                            y = "Percentage of unmapped reads (%)") 

reads_spike_num <- ggplot(anno,
                   aes(x = mapped, y = spike_percentage * 100,
                       col = as.factor(chip_id), 
                       label = as.character(cell_number), 
                       height = 600, width = 2000)) +
                   geom_text(fontface = 3, alpha = 0.5) + 
                   geom_vline(xintercept = cut_off_reads, 
                              colour="grey", linetype = "longdash") +
                   geom_hline(yintercept = cut_off_spike * 100, 
                              colour="grey", linetype = "longdash") +
                   labs(x = "Total mapped reads / sample",
                        y = "Percentage of spike-in reads (%)") 

plot_grid(genes_unmapped + theme(legend.position=c(.7,.9)), 
          genes_spike + theme(legend.position = "none"),
          labels = letters[1:2])
plot_grid(reads_unmapped_num + theme(legend.position = "none"), 
          reads_spike_num + theme(legend.position = "none"),
          labels = letters[3:4])
```

## PCA

### Before filter
Select the most variable human genes
```{r select-gene-before}
## look at human genes
eset_hs <- eset[fData(eset)$source == "H. sapiens", ]
head(featureNames(eset_hs))

## remove genes of all 0s
eset_hs_clean <- eset_hs[rowSums(exprs(eset_hs)) != 0, ]
dim(eset_hs_clean)

## convert to log2 cpm
mol_hs_cpm <- cpm(exprs(eset_hs_clean), log = TRUE)
hist(rowMeans(mol_hs_cpm))

## variable genes
mol_hs_cpm_var <- log(apply(mol_hs_cpm, 1, var))
hist(mol_hs_cpm_var)
sum(mol_hs_cpm_var > 0.5)
```

Using the `r sum(mol_hs_cpm_var > 0.5)` most variable genes to perform PCA

```{r pca-before}
## pca of top variable genes
mol_hs_cpm <- mol_hs_cpm[mol_hs_cpm_var > 0.5, ]
pca_hs <- run_pca(mol_hs_cpm) 
plot_pca(pca_hs$PCs, pcx = 1, pcy = 2, explained = pca_hs$explained,
         metadata = pData(eset_hs_clean), color = "experiment")

## combine to investigate the effect
pca_anno <- cbind(anno, pca_hs$PCs)

## total mapped vs pc1
pc1_reads <- ggplot(pca_anno, aes(x = mapped, y = PC1)) +
                     geom_text(aes(label = cell_number, 
                                   col = filter_all, alpha = 0.5)) + 
                     scale_colour_manual(values=cbPalette) +
                     geom_smooth()

## unmapped ratio vs pc1
pc1_unmapped <- ggplot(pca_anno, aes(x = unmapped_ratios, y = PC1)) +
                       geom_text(aes(label = cell_number,
                                     col = filter_all, alpha = 0.5)) + 
                       scale_colour_manual(values=cbPalette) + 
                       geom_smooth()

## spike-in ratio vs pc1
pc1_spike <- ggplot(pca_anno, aes(x = spike_percentage, y = PC1)) +
                     geom_text(aes(label = cell_number,
                                   col = filter_all, alpha = 0.5)) + 
                     scale_colour_manual(values=cbPalette) + 
                     geom_smooth()

## number of detected gene vs pc1
pc1_gene <- ggplot(pca_anno, aes(x = detect_hs, y = PC1)) +
                     geom_text(aes(label = cell_number,
                                   col = filter_all, alpha = 0.5)) + 
                     scale_colour_manual(values=cbPalette) + 
                     geom_smooth()

plot_grid(pc1_reads + theme(legend.position=c(.7,.5)),
          pc1_unmapped + theme(legend.position = "none"),
          pc1_spike + theme(legend.position = "none"), 
          pc1_gene + theme(legend.position = "none"),
          labels = LETTERS[1:4])
```

### After filter
```{r pca-after}
## filter bad cells
eset_hs_clean_filter <- eset_hs_clean[,anno$filter_all]
dim(eset_hs_clean_filter)

## convert to log2 cpm
mol_hs_cpm_filter <- cpm(exprs(eset_hs_clean_filter), log = TRUE)
hist(rowMeans(mol_hs_cpm_filter))
stopifnot(rownames(anno[anno$filter_all,]) == colnames(mol_hs_cpm_filter))

## variable genes
mol_hs_cpm_filter_var <- log(apply(mol_hs_cpm_filter, 1, var))
hist(mol_hs_cpm_filter_var)
sum(mol_hs_cpm_filter_var > 0.5)

## pca of top variable genes
mol_hs_cpm_filter<- mol_hs_cpm_filter[mol_hs_cpm_filter_var > 0.5, ]
pca_hs_filter <- run_pca(mol_hs_cpm_filter) 
plot_pca(pca_hs_filter$PCs, pcx = 1, pcy = 2, explained = pca_hs_filter$explained,
         metadata = pData(eset_hs_clean_filter), color = "experiment")

## combine to investigate the effect
anno_filter <- anno[anno$filter_all,]
pca_anno_filter <- cbind(anno_filter, pca_hs_filter$PCs)

## total mapped vs pc1
pc1_reads_filter <- ggplot(pca_anno_filter, aes(x = mapped, y = PC1)) +
                     geom_text(aes(label = cell_number, 
                                   alpha = 0.5)) + 
                     scale_colour_manual(values=cbPalette) +
                     geom_smooth()

## unmapped ratio vs pc1
pc1_unmapped_filter <- ggplot(pca_anno_filter, aes(x = unmapped_ratios, y = PC1)) +
                       geom_text(aes(label = cell_number,
                                     alpha = 0.5)) + 
                       scale_colour_manual(values=cbPalette) + 
                       geom_smooth()

## spike-in ratio vs pc1
pc1_spike_filter <- ggplot(pca_anno_filter, aes(x = spike_percentage, y = PC1)) +
                     geom_text(aes(label = cell_number,
                                   alpha = 0.5)) + 
                     scale_colour_manual(values=cbPalette) + 
                     geom_smooth()

## number of detected gene vs pc1
pc1_gene_filter <- ggplot(pca_anno_filter, aes(x = detect_hs, y = PC1)) +
                     geom_text(aes(label = cell_number,
                                   alpha = 0.5)) + 
                     scale_colour_manual(values=cbPalette) + 
                     geom_smooth()

plot_grid(pc1_reads_filter + theme(legend.position=c(.7,.5)),
          pc1_unmapped_filter + theme(legend.position = "none"),
          pc1_spike_filter + theme(legend.position = "none"), 
          pc1_gene_filter + theme(legend.position = "none"),
          labels = LETTERS[1:4])
```
```

