---
title: "Analysis of totals counts"
author: "John Blischak"
date: 2017-08-14
output: workflowr::wflow_html
---









This analysis explores the total counts of reads and molecules mapped to each
source (human, fly, worm, and ERCC).

## Setup

```{r packages, message=FALSE}
library("cowplot")
library("dplyr")
library("ggplot2")
library("tidyr")
theme_set(theme_cowplot())
source("../code/functions.R")
library("Biobase")
```

Import data.

```{r import}
eset <- readRDS("../data/eset.rds")
anno <- pData(eset)
```

Remove samples with bad cell number or TRA-1-60.

```{r filter}
eset_qc <- eset[, anno$cell_number == 1 & anno$tra1.60]
anno_qc <- pData(eset_qc)
```

Extract the columns relevant for this analysis.

```{r totals}
totals <- anno_qc %>% select(experiment:well, raw:mol_hs)
```

## What percentage of reads are mapped to the genome?

```{r percent-mapped}
totals_long <- totals %>%
  gather(key = "type", value = "count", raw:mol_hs) %>%
  filter(type %in% c("raw", "umi", "mapped", "molecules")) %>%
  mutate(type = factor(type, levels = c("raw", "umi", "mapped", "molecules"),
                       labels = c("Raw reads", "Reads with UMI",
                                  "Reads mapped\nto genome",
                                  "Molecules mapped\nto genome")))
head(totals_long)
ggplot(totals_long, aes(x = type, y = count)) +
  geom_boxplot() +
  labs(x = "Classification", y = "Number of sequences",
       title = "Processing of raw reads")
```

Category  | mean | sd | \% of raw (mean)
------------- | ------------- | ------------- | -------------
Raw reads  | `r mean(totals$raw)` | `r sd(totals$raw)` | `r specify_decimal(mean(totals$raw / totals$raw), k = 2)`
Reads with UMI  | `r mean(totals$umi)` | `r sd(totals$umi)` | `r specify_decimal(mean(totals$umi / totals$raw), k = 2)`
Reads mapped to genome  | `r mean(totals$mapped)` | `r sd(totals$mapped)` | `r specify_decimal(mean(totals$mapped / totals$raw), k = 2)`
Molecules mapped to genome  | `r mean(totals$molecules)` | `r sd(totals$molecules)` | `r specify_decimal(mean(totals$molecules / totals$raw), k = 2)`

## How does the number of mapped reads vary by C1 chip?

```{r mapped-per-chip}
mapped <- totals_long %>%
  filter(type == "Reads mapped\nto genome")
ggplot(mapped, aes(x = experiment, y = count)) +
  geom_boxplot() +
  labs(x = "C1 chip", y = "Number of mapped reads",
       title = "Mapped reads per C1 chip") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

```{r total-mapped-per-chip}
mapped_sum <- mapped %>%
  group_by(experiment) %>%
  summarize(total = sum(count))
ggplot(mapped_sum, aes(x = experiment, y = total)) +
  geom_bar(stat= "identity") +
  labs(x = "C1 chip", y = "Total number of mapped reads",
       title = "Total mapped reads per C1 chip") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

Sorted by total number of mapped reads (in millions).

```{r total-mapped-table}
mapped_sum %>%
  arrange(total) %>%
  mutate(total = total / 10^6) %>%
  knitr::kable()
```

## What is the conversion of reads to molecules?

```{r reads-to-molecules, fig.width=10, fig.height=10}
convert_all <- ggplot(totals, aes(x = mapped, y = molecules)) +
  geom_point() +
  labs(x = "Number of reads mapped to any genome",
       y = "Number of molecules mapped to any genome",
       title = "Conversion of reads to molecules")
convert_hs <- ggplot(totals, aes(x = reads_hs, y = mol_hs)) +
  geom_point() +
  labs(x = "Number of reads mapped to human genome",
       y = "Number of molecules mapped to human genome",
       title = "Conversion of reads to molecules for human")
convert_dm <- ggplot(totals, aes(x = reads_dm, y = mol_dm)) +
  geom_point() +
  labs(x = "Number of reads mapped to fly genome",
       y = "Number of molecules mapped to fly genome",
       title = "Conversion of reads to molecules for fly")
convert_ce <- ggplot(totals, aes(x = reads_ce, y = mol_ce)) +
  geom_point() +
  labs(x = "Number of reads mapped to worm genome",
       y = "Number of molecules mapped to worm genome",
       title = "Conversion of reads to molecules for worm")

plot_grid(convert_all, convert_hs, convert_dm, convert_ce,
          labels = letters[1:4])
```

## Drosophila spike-in

```{r totals-fly-perc}
p_reads_dm <- ggplot(anno_qc, aes(x = as.factor(fly), y = reads_dm / mapped * 100)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Percentage of mapped reads that mapped to fly genome")
p_mol_dm <- ggplot(anno_qc, aes(x = as.factor(fly), y = mol_dm / molecules * 100)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Percentage of mapped molecules that mapped to fly genome")
plot_grid(p_reads_dm, p_mol_dm, labels = letters[1:2])
```

```{r totals-fly-counts}
p_reads_dm_counts <- ggplot(anno_qc, aes(x = as.factor(fly), y = reads_dm)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Number of reads that mapped to fly genome")
p_mol_dm_counts <- ggplot(anno_qc, aes(x = as.factor(fly), y = mol_dm)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Number of molecules that mapped to fly genome")
plot_grid(p_reads_dm_counts, p_mol_dm_counts, labels = letters[1:2])
```

```{r read-classification-per-sample}
totals_percent <- anno_qc %>%
  mutate(sample = paste(experiment, well, sep = "-"),
         sample = factor(sample, levels = sample[order(reads_dm / raw)])) %>%
  arrange(reads_dm / raw) %>%
  group_by(experiment, sample, fly, worm, ERCC) %>%
  summarize(p_reads_ce = reads_ce / raw * 100,
            p_reads_dm = reads_dm / raw * 100,
            p_reads_ercc = reads_ercc / raw * 100,
            p_reads_hs = reads_hs / raw * 100,
            p_unmapped = unmapped / raw * 100,
            p_no_umi = (raw - unmapped - mapped) / raw * 100) %>%
  gather(key = "type", value = "percent", p_reads_ce:p_no_umi) %>%
  ungroup() %>%
  mutate(type = factor(type, levels = c("p_no_umi", "p_unmapped", "p_reads_hs",
                                        "p_reads_ercc", "p_reads_ce", "p_reads_dm"),
                       labels = c("No UMI", "Unmapped", "H. sapiens", "ERCC",
                                  "C. elegans", "D. melanogaster")),
         fly = factor(fly, levels = c(5000, 50000), labels = c("5 pg fly spike-on",
                                                               "50 pg fly spike-on")))

ggplot(totals_percent, aes(x = sample, y = percent, fill = type)) +
  geom_col() +
  facet_wrap(~fly) +
  labs(x = "Sample", y = "Percent", title = "Read classification per sample") +
  theme(legend.title = element_blank())
```

## C. elegans spike-in

```{r totals-worm-perc}
p_reads_ce <- ggplot(anno_qc, aes(x = as.factor(worm), y = reads_ce / mapped * 100)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Percentage of mapped reads that mapped to worm genome")
p_mol_ce <- ggplot(anno_qc, aes(x = as.factor(worm), y = mol_ce / molecules * 100)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Percentage of mapped molecules that mapped to worm genome")
plot_grid(p_reads_ce, p_mol_ce, labels = letters[1:2])
```

```{r totals-worm-counts}
p_reads_ce_counts <- ggplot(anno_qc, aes(x = as.factor(worm), y = reads_ce)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Number of mapped reads that mapped to worm genome")
p_mol_ce_counts <- ggplot(anno_qc, aes(x = as.factor(worm), y = mol_ce)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Number of mapped molecules that mapped to worm genome")
plot_grid(p_reads_ce_counts, p_mol_ce_counts, labels = letters[1:2])
```


## ERCC spike-in

```{r totals-ercc-perc}
p_reads_ercc <- ggplot(anno_qc, aes(x = as.factor(ERCC), y = reads_ercc / mapped * 100)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Percentage of mapped reads that mapped to ERCCs")
p_mol_ercc <- ggplot(anno_qc, aes(x = as.factor(ERCC), y = mol_ercc / molecules * 100)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Percentage of mapped molecules that mapped to ERCCs")
plot_grid(p_reads_ercc, p_mol_ercc, labels = letters[1:2])
```

```{r totals-ercc-counts}
p_reads_ercc_counts <- ggplot(anno_qc, aes(x = as.factor(ERCC), y = reads_ercc)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Number of reads that mapped to ERCCs")
p_mol_ercc_counts <- ggplot(anno_qc, aes(x = as.factor(ERCC), y = mol_ercc)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Number of molecules that mapped to ERCCs")
plot_grid(p_reads_ercc_counts, p_mol_ercc_counts, labels = letters[1:2])
```



