---
title: "Analysis of totals counts"
author: "John Blischak"
date: 2017-08-14
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

This analysis explores the total counts of reads and molecules mapped to each
source (human, fly, worm, and ERCC).

## Setup

```{r packages, message=FALSE}
library("cowplot")
library("dplyr")
library("ggplot2")
library("readr")
library("stringr")
library("tidyr")
theme_set(theme_cowplot())
source("../code/functions.R")
```

Import totals counts.

```{r read-totals, message=FALSE}
totals <- read_tsv("../data/totals.txt")
dim(totals)
totals[1:5, 1:10]
colnames(totals)
```

Import annotation.

```{r read-anno, message=FALSE}
anno <- read_tsv("../data/batch1_qc.txt",
                 col_types = cols(ERCC = col_character()))
# Cleanup ERCC column
anno$ERCC <- ifelse(is.na(anno$ERCC), "Not added", anno$ERCC)
anno$ERCC <- factor(anno$ERCC, levels = c("Not added", "1:100000", "1:50000"),
                    labels = c("Not added", "100x dilution", "50x dilution"))
dim(anno)
colnames(anno)
```

Merge the data sets.

```{r merge}
totals_anno <- merge(totals, anno, by = c("experiment", "well"))
stopifnot(nrow(totals_anno) == nrow(totals))
```

Remove samples with bad cell number or TRA-1-60.

```{r filter}
totals_anno <- totals_anno %>%
  filter(cell_number == 1, tra1.60 == 1)
dim(totals_anno)
```

## What percentage of reads are mapped to the genome?

```{r percent-mapped}
totals_long <- totals %>%
  gather(key = "type", value = "count", raw:mol_hs) %>%
  filter(type %in% c("raw", "umi", "mapped", "molecules")) %>%
  mutate(type = factor(type, levels = c("raw", "umi", "mapped", "molecules"),
                       labels = c("Raw reads", "Reads with UMI",
                                  "Reads mapped\nto genome",
                                  "Molecules mapped\nto genome")))
head(totals_long)
ggplot(totals_long, aes(x = type, y = count)) +
  geom_boxplot() +
  labs(x = "Classification", y = "Number of sequences",
       title = "Processing of raw reads")
```

Category  | mean | sd | \% of raw (mean)
------------- | ------------- | ------------- | -------------
Raw reads  | `r mean(totals$raw)` | `r sd(totals$raw)` | `r specify_decimal(mean(totals$raw / totals$raw), k = 2)`
Reads with UMI  | `r mean(totals$umi)` | `r sd(totals$umi)` | `r specify_decimal(mean(totals$umi / totals$raw), k = 2)`
Reads mapped to genome  | `r mean(totals$mapped)` | `r sd(totals$mapped)` | `r specify_decimal(mean(totals$mapped / totals$raw), k = 2)`
Molecules mapped to genome  | `r mean(totals$molecules)` | `r sd(totals$molecules)` | `r specify_decimal(mean(totals$molecules / totals$raw), k = 2)`

## What is the conversion of reads to molecules?

```{r reads-to-molecules, fig.width=10, fig.height=10}
convert_all <- ggplot(totals, aes(x = mapped, y = molecules)) +
  geom_point() +
  labs(x = "Number of reads mapped to any genome",
       y = "Number of molecules mapped to any genome",
       title = "Conversion of reads to molecules")
convert_hs <- ggplot(totals, aes(x = reads_hs, y = mol_hs)) +
  geom_point() +
  labs(x = "Number of reads mapped to human genome",
       y = "Number of molecules mapped to human genome",
       title = "Conversion of reads to molecules for human")
convert_dm <- ggplot(totals, aes(x = reads_dm, y = mol_dm)) +
  geom_point() +
  labs(x = "Number of reads mapped to fly genome",
       y = "Number of molecules mapped to fly genome",
       title = "Conversion of reads to molecules for fly")
convert_ce <- ggplot(totals, aes(x = reads_ce, y = mol_ce)) +
  geom_point() +
  labs(x = "Number of reads mapped to worm genome",
       y = "Number of molecules mapped to worm genome",
       title = "Conversion of reads to molecules for worm")

plot_grid(convert_all, convert_hs, convert_dm, convert_ce,
          labels = letters[1:4])
```

## Drosophila spike-in

```{r totals-fly-perc}
p_reads_dm <- ggplot(totals_anno, aes(x = as.factor(fly), y = reads_dm / mapped * 100)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Percentage of mapped reads that mapped to fly genome")
p_mol_dm <- ggplot(totals_anno, aes(x = as.factor(fly), y = mol_dm / molecules * 100)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Percentage of mapped molecules that mapped to fly genome")
plot_grid(p_reads_dm, p_mol_dm, labels = letters[1:2])
```

```{r totals-fly-counts}
p_reads_dm_counts <- ggplot(totals_anno, aes(x = as.factor(fly), y = reads_dm)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Number of reads that mapped to fly genome")
p_mol_dm_counts <- ggplot(totals_anno, aes(x = as.factor(fly), y = mol_dm)) +
  geom_boxplot() +
  labs(x = "Fly spike-in concentration",
       y = "Number of molecules that mapped to fly genome")
plot_grid(p_reads_dm_counts, p_mol_dm_counts, labels = letters[1:2])
```

## C. elegans spike-in

```{r totals-worm-perc}
p_reads_ce <- ggplot(totals_anno, aes(x = as.factor(worm), y = reads_ce / mapped * 100)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Percentage of mapped reads that mapped to worm genome")
p_mol_ce <- ggplot(totals_anno, aes(x = as.factor(worm), y = mol_ce / molecules * 100)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Percentage of mapped molecules that mapped to worm genome")
plot_grid(p_reads_ce, p_mol_ce, labels = letters[1:2])
```

```{r totals-worm-counts}
p_reads_ce_counts <- ggplot(totals_anno, aes(x = as.factor(worm), y = reads_ce)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Percentage of mapped reads that mapped to worm genome")
p_mol_ce_counts <- ggplot(totals_anno, aes(x = as.factor(worm), y = mol_ce)) +
  geom_boxplot() +
  labs(x = "Worm spike-in concentration",
       y = "Percentage of mapped molecules that mapped to worm genome")
plot_grid(p_reads_ce_counts, p_mol_ce_counts, labels = letters[1:2])
```


## ERCC spike-in

```{r totals-ercc-perc}
p_reads_ercc <- ggplot(totals_anno, aes(x = as.factor(ERCC), y = reads_ercc / mapped * 100)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Percentage of mapped reads that mapped to ERCCs")
p_mol_ercc <- ggplot(totals_anno, aes(x = as.factor(ERCC), y = mol_ercc / molecules * 100)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Percentage of mapped molecules that mapped to ERCCs")
plot_grid(p_reads_ercc, p_mol_ercc, labels = letters[1:2])
```

```{r totals-ercc-counts}
p_reads_ercc_counts <- ggplot(totals_anno, aes(x = as.factor(ERCC), y = reads_ercc)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Number of reads that mapped to ERCCs")
p_mol_ercc_counts <- ggplot(totals_anno, aes(x = as.factor(ERCC), y = mol_ercc)) +
  geom_boxplot() +
  labs(x = "ERCC spike-in concentration",
       y = "Number of molecules that mapped to ERCCs")
plot_grid(p_reads_ercc_counts, p_mol_ercc_counts, labels = letters[1:2])
```

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
